Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-31 tagless/tagless-final.tex
A little bit more work on the introduction

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}

@@

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a Boolean} \justifies b:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
datatype (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
contributions.  No matter how we represent the object language in the
metalanguage, the representation can be performed by hand or more
conveniently by a parser or preprocessor such as camlp4.  We leave aside
the issue of whether to program a parser or preprocessor using dependent
types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

let rec eval0 env = function 
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}
If our OCaml-like metalanguage were untyped, the code above would be acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \cite[\S1.1--2]{WalidICFP02}.

If we use OCaml or some other typed language as the meta-language, 
the function |eval0| is ill-typed. The line |B b|
says that |eval0| returns a boolean, whereas the next line says
the result is a function. All branches of a pattern-match form must
yield values of the same type, however. 
A related problem is the type of the environment |env|: since variables
in the language may be either of boolean or function types, a regular
OCaml list cannot be used for holding the corresponding values.

The usual solution is to introduce a Universal type (section 1.3 of
\cite{WalidICFP02}) which contains both of these:

\begin{code}
  type u = UB of bool | UA of (u -> u)
\end{code}

We can then write an interpreter which is accepted by the typecheker
\begin{code}
let rec eval env = function
  | V VZ -> List.hd env
  | V (VS v) -> eval (List.tl env) (V v)
  | B b -> UB b
  | L e -> UA (fun x -> eval (x::env) e)
  | A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
and whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term:

\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}

Unfortunately, the result is tagged: |UB true| rather than just
|true|. We also note that the eval function is partial in two
respects: first there is an inexhaustive pattern-matching 
|match ... with UA f -> ...| in the 
line |A (e1,e2)|.
Second, the occurrences of |List.hd| and |List.tl|.
The failure of the latter corresponds to the evaluation of `open'
terms. One can see that if all our terms are closed, |List.hd| and
|List.tl| above will never be applied to the empty list. The partial
pattern match in the |A (e1,e2)| line raises an exception when we try to
evaluate something like |A (B true, B false)|, that is when the first
operand is not a function.

Suppose our object language were typed too. Then terms like
|A (B true, B false)| would be ill-typed. Thus, if our evaluator
accepted only typed terms, the missing alternatives in the |A|
line would not occur. Alas, what is obvious to us is not obvious to
the meta-language. There is still a pattern match, and we still have
to attach and eliminate the UB and UA tags, even when there is no
longer a need for them. See \cite{WalidICFP02}, Section 1.4 for more details.

In short: It is not satisfactory
to use a Universal type because tagging hurts performance and
necessitates partial pattern matching in the metalanguage, even for a
well-typed object term.

The universal type fails to express in the metalanguage
that the object program is well-typed.  This failure gives rise to
tagging and untagging (that is, apparently inexhaustive pattern\hyp
matching) operations that incur a performance penalty in interpretation
and impair optimality in partial evaluation.  In short, the
universal\hyp type solution is not type\hyp preserving.

\subsection{Other solutions}

 The problem of typed direct
  intrepreter of a typed language is difficult and thought to require
  GADT or dependent types. 
  In fact, several papers introducing GADTs (Xi03 and SPJ) as well as 
  (language in Walid03) and Concoqtion, written 2003-2007, used the 
  typed tagless intrepreter as the motivating example.
  Taha (in Jones optimality paper, I think) said that typed
  interpreter of the typed language was not solved then.
  Or, should we refer to Sec `Other solutions'.

But how should one design a type system to type
terms like |test1|? GADTs seem necessary. Similarly, how does one design a
typesystem that can keep track of the open/closed status of terms, so that we
can statically insure that only closed terms are passed to |eval| and that
the empty list checking in |List.hd| and |List.tl| can be eliminated.
The latter is an even more interesting problem requiring even
more advanced type systems 
(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
\jacques{I hope that is the right Davies paper}

This is also the main motivating problem (albeit in the context of
a first-order language) for the introduction of GADTs 
into Haskell \cite{PeytonJonesICFP2006}.  This was also a motivating
factor behind the work of Xi on GADTs \cite{XiGADTs}, and some of the
recent work on dependent types \cite{WalidICFP02,WalidConcoqtion}.

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.
\oleg{I feel we need to give some details of the other solutions here, 
to give the reader the feeling of
how advanced solutions were proposed and how rarely they are
implemented.}
\jacques{I have rearranged this section so that it should be clear how
to extend it to explain these other solutions}

\subsection{The ``Final'' Proposal}\label{ourapproach}

@@

We have realized that if we change the encoding of the term from data
constructors to functions, we can encode a tagless interpreter for a
typed object language into a metalanguage with a very simple type system.

@@

This time we use higher-order abstract syntax \cite{DaleMiller}
rather than de Bruijn indices.  The
language is close to the language of \cite{XiPOPL03}, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Fibonacci, factorial, and power.

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}

@@

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a boolean} \justifies b:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
datatype (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  No matter how we represent the object language in the
metalanguage, the representation can be performed by hand or more
conveniently by a parser or preprocessor such as camlp4.  We leave aside
the issue of whether to program a parser or preprocessor using dependent
types \citep{WalidICFP02} or not \citep{baars-typing}.

@@

let rec eval0 env = function 
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

If our OCaml-like metalanguage were untyped, the code above would be acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \cite[\S1.1--2]{WalidICFP02}.

However, the function |eval0| is ill-typed
if we use OCaml or some other typed language as the metalanguage.
The line |B b|
says that |eval0| returns a boolean, whereas the next line |L e| says
the result is a function, but all branches of a pattern-match form must
yield values of the same type. 
A related problem is the type of the environment |env|: a regular
OCaml list cannot hold both boolean and function values.

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions.
\begin{code}
type u = UB of bool | UA of (u -> u) | ...
\end{code}
We can then write a typed interpreter
\begin{code}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f ->
               f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term.
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}

The unfortunate tag |UB| in the result reflects that |eval| is a partial
function, in two ways.  First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term.
\begin{code}
let test3 = V VZ
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

If we evaluate only closed terms that have been type-checked, then
|eval| would never fail. Alas, this soundness is not obvious to the
metalanguage, whose type system we must still appease with the
nonexhaustive pattern matching in |lookup| and |eval| and the tags |UB|
and |UA| \citep[\S1.4]{WalidICFP02}.  In other words, the algebraic data
types above fail to express in the metalanguage that the object program
is well-typed.  This failure necessitates tagging and nonexhaustive
pattern\hyp matching operations that incur a performance penalty in
interpretation and impair optimality in partial evaluation
\citep{taha-tag}.  In short, the universal\hyp type solution is
unsatisfactory because it does not preserve typing.

\subsection{Solutions Using Fancier Types}

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \citep{xi-guarded,peyton-jones-simple} and
on dependent types \citep{WalidICFP02,fogarty-concoqtion}.

For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms like |test3|
\citep{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}; here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed.

\subsection{Our Final Proposal}\label{ourapproach}

If we change the encoding of the term from data
constructors to functions, we can encode a tagless interpreter for a
typed object language into a metalanguage with a very simple type system.

@@

@@

This time we use higher-order abstract syntax \cite{DaleMiller}
rather than de Bruijn indices.  The
language is close to the language of \citet{xi-guarded}, without the polymorphic
lift but with more constants so we can write better examples, enough to
express Fibonacci, factorial, and power.
